Appendix 4: Results

### 2016-02-11 Harris_2015/manuscript-materials/draft.Rmd
editor comments

**Abstract:** Inferring species interactions from co-occurrence data is one of
the most controversial tasks in community ecology. One difficulty is that a
single pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, the negative correlation between
two competing species can be reversed in the presence of a third species that is
capable of outcompeting both of them. Here, I apply models from statistical
physics, called Markov networks or Markov random fields, that can predict
the direct and indirect consequences of any possible species interaction matrix.
Interactions in these models can also be estimated from observed co-occurrence
rates via maximum likelihood, controlling for indirect effects. Using simulated
landscapes with known pairwise interaction strengths, I evaluated Markov
networks and six existing approaches. The Markov networks consistently
outperformed other methods, correctly isolating direct interactions between
species pairs even when indirect interactions or abiotic factors largely
overpowered them. Two computationally efficient approximations, based on
controlling for indirect effects with linear or generalized linear models, also
performed well. Indirect effects reliably caused a common null modeling approach
to produce incorrect inferences, however.

@@

\noindent
To the extent that nontrophic species interactions (such as competition) affect
community assembly, ecologists might expect to find signatures of these
interactions in species composition data [@macarthur_population_1958;
@diamond_island_1975]. Despite decades of work and several major controversies,
however [@lewin_santa_1983; @strong_ecological_1984; @connor_checkered_2013],
existing methods for detecting competition's effects on community structure are
unreliable [@gotelli_empirical_2009]. In particular, species' effects on one
another can become lost in a web of indirect effects. For example, the
competitive interaction between the two shrub species in Figure 1A is
obscured by their shared tendency to occur in unshaded areas (Figure 1B). While
ecologists have long known that indirect effects can overwhelm direct ones at
the landscape level [@levine_competitive_1976], the
vast majority of our methods for drawing inferences from observational data do
not control for these effects [e.g. @diamond_island_1975;
@strong_ecological_1984; @gotelli_empirical_2009; @veech_probabilistic_2013;
@pollock_understanding_2014]. To the extent that indirect interactions like
those in Figure 1 are generally important, existing
methods will not provide much evidence regarding species interactions.

While competition doesn't reliably reduce co-occurrence rates at the
whole-landscape level (as most methods assume), it does still leave a
signal in the data (Figure 1C). For example, after partitioning the data set
into shaded and unshaded sites, there will be co-occurrence deficits in
each subset that wouldn't otherwise be apparent. More generally, controlling for
other species in the network will often be important for obtaining reliable
estimates of direct (conditional, or all-else-equal) effects.  This kind of
precision is difficult to obtain from null models, which only test the most
extreme possible hypothesis: that *all* direct and indirect interactions are
exactly zero. Nevertheless, null models have dominated this field for more than
three decades [@strong_ecological_1984; @gotelli_empirical_2009].

Following @azaele_inferring_2010, this paper shows that Markov networks
[undirected graphical models also known as Markov random fields;
@murphy_machine_2012] can provide a framework for understanding the
landscape-level consequences of pairwise species interactions,
and for estimating them from observed presence-absence matrices. Markov networks
have been used in many scientific fields in similar contexts for decades, from
physics [where nearby particles interact magnetically; @cipra_introduction_1987]
to spatial statistics [where adjacent grid cells have correlated values;
@harris_contact_1974; @gelfand_modelling_2005]. While community ecologists
explored some related approaches in the 1980's [@whittam_species_1981],
they used severe approximations that led to unintelligible results
[e.g. "probabilities" greater than one; @gilpin_factors_1982].

Below, I demonstrate Markov networks' ability to produce exact predictions
about the direct and indirect consequences of an interaction matrix, and also to
make inferences about the species interactions that contributed to an observed
set of co-occurrences. Using simulated data sets where the "true" interactions
are known, I compare this approach with several existing methods. Finally, I
discuss opportunities for extending the approach presented here to other
problems in community ecology, e.g. quantifying the overall effect of species
interactions on occurrence rates [@roughgarden_competition_1983] and
disentangling the effects of biotic versus abiotic interactions on species
composition [@pollock_understanding_2014].

@@

\noindent where the second sum is over all $\frac{1}{2}n(n-1)$ pairs of $n$
species. In this model, $\alpha_{i}$ is an intercept term determining the amount
that the presence of species $i$ contributes to the log-probability of
$\vec{y}$; it directly controls the prevalence of species $i$. Similarly,
$\beta_{ij}$ is the amount that the co-occurrence of species $i$ and species $j$
contributes to the log-probability; it controls the conditional relationship
between two species, i.e. the probability that they will be found together,
after controlling for the other species in the network (Figure 2A, Figure 2B).
For example, if $\beta_{ij} = +2$, then each species' odds of occurrence would be
$e^2$ times higher when the other one is present (as compared with otherwise
equivalent sites). The relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur. As a result, the model tends---all else
equal---to produce assemblages where many positively-associated species pairs
co-occur and few negatively-associated pairs do (just as an ecologist might
expect). When all else is *not* equal (e.g. Figure 1, where the presence of one
competitor is associated with release from another competitor), then predicting
species' overall co-occurrence rates can be more complicated, and may require
summing over the different possible assemblages (Figure 2B).

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a set of independent $\vec{y}$ vectors
indicating which species are present at each site on the landscape). When the
number of species is reasonably small (less than about 30), one can find
exact maximum likelihood estimates for all of the $\alpha$ and $\beta$
coefficients given a presence-absence matrix by numerically
optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
ensuring that this procedure will converge on the global maximum. I used the
rosalia package [@harris_rosalia_2015] for the R programming language
[@r_core_team_r_2015] to calculate $p(\vec{y}; \alpha, \beta)$ and its
gradients [@murphy_machine_2012]; the package passes these functions to the
"BFGS" method in R's general-purpose optimizer, which then estimates the Markov
network parameters.

@@

I then generated landscapes with up to 20 interacting species at 25, 200, or
1600 sites using three increasingly complex models (50 replicates for each
combination of size and model; see Appendix 2 for details). I randomly drew the
“true” coefficient magnitudes for each replicate landscape from exponential
distributions so that most species pairs interacted negligibly but a few pairs
interacted strongly enough that their effects could propagate indirectly to
other species in the network.

The first set of 20-species landscapes, like the landscapes with three species,
were generated directly from a Markov network to ensure that the model could
recover the parameters used to generate the "observed" co-occurrence data.
Then, I added two environmental factors that varied from location to location
across the simulated landscapes, and simulated a new set of co-occurrence data
so that species' $\alpha$ coefficients depended on the local environment.
The latter set of simulated landscapes provide an important test of the methods’
ability to distinguish co-occurrence patterns that were generated from pairwise
biotic interactions from those that were generated by external forces like
abiotic environmental filtering. This task was made especially difficult
because---as with most analyses of presence-absence data for co-occurrence
patterns---the inference procedure did not have access to any information about
the environmental or spatial variables that helped shape the landscape [cf
@connor_checkered_2013].  I generated the final set of landscapes with an
abundance-based model that included per-capita interaction rates instead of
per-species interaction rates.

@@

I also evaluated six alternative methods: five from the existing literature,
plus a novel combination of two of these methods. The first alternative interaction
metric was the sample correlation between species' presence-absence vectors,
which summarizes their marginal association.  Next, I used partial correlations,
which summarize species' conditional relationships. This approach is
common in molecular biology [@friedman_sparse_2008], but is rare in ecology
(see @albrecht_spatial_2001 and @faisal_inferring_2010 for two exceptions).
In the context of non-Gaussian data, the partial correlation can be thought of
as a computationally efficient approximation to the full Markov network model
[@loh_structure_2013]. Because partial correlations are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate provided by the corpcor package’s `pcor.shrink` function with the
default settings [@schafer_corpcor_2014].

@@

The next method, described in @gotelli_empirical_2009, involved
simulating new landscapes from a null model that retains the row and column
sums of the original matrix [@strong_ecological_1984]. I used the $Z$-scores
computed by the Pairs software described in @gotelli_empirical_2009 as my null
model-based estimator of species interactions.

@@

For the larger landscapes, I evaluated the relationship between each method's
estimates and the "true" interaction strengths. To ensure that the different
test statistics (e.g. correlations versus $Z$ scores) were on a common scale, I
rescaled them using linear regression through the origin. I then calculated the
proportion of variance explained for different combinations of model type and
landscape size (compared with a null model that assumed all interaction
strengths to be zero).

@@

\noindent
\textbf{Twenty species.}
In general, each model's performance was highest for large landscapes with
simple assembly rules and no environmental heterogeneity (Figure 3).
Despite some variability across contexts, the rank ordering across methods was
very consistent.  In particular, the four methods that controlled for indirect
effects (the Markov network, the generalized linear models, and the two partial
correlation-based methods) always matched or outperformed those that did not.
The Markov network consistently performed best of all. As anticipated by
@lee_learning_2012, generalized linear models closely
approximated the Markov network estimates (Figure 4A), especially when the data
sets were very large (Figure 3). As reviewed in @gotelli_empirical_2009,
however, most analyses in this field of ecology involve fewer than 50 sites,
and the gap between the Markov network and GLMs is larger in this context. As
shown in Appendix 4, the standard errors associated with the estimates in Figure
3 are small (less than 0.01), so the differences among methods should not be
attributed to sampling error.

Of the methods that did not control for indirect effects, Figure 3 shows that
simple correlation coefficients provided a more reliable indicator of species'
true interaction strengths than either the joint species distribution model
(BayesComm) or the null model (Pairs). The estimates from these approaches
were tightly correlated (after controlling for the size of the landscape)
suggesting that the null model only contains a noisy version of the
same information that could be obtained more easily and interpretably with
simple correlation coefficients (Figure 4B).

Finally, we can evaluate the models' statistical inferences (focusing on the
first two simulation types, for which the true interaction rates are easiest to
interpret).  The Markov network's approximate Type I error rate (defined here
as the probability that 0 fell outside the 95% confidence interval for a pair of
species where $|\beta_{ij}|<0.1$) depended on the simulation type: 0.02 for
simulations that matched the model's assumptions, versus 0.14 for simulations
that included environmental heterogeneity (see Appendix 4 for confidence
interval coverage across a range of $\beta_{ij}$ values). The null model's
Type I error rates were 0.30 and 0.50 for the constant and heterogeneous
landscapes, respectively---far higher than the nominal 0.05 rate. Figure
4C shows, across a range of true interaction strengths, the probability that
the null model or the Markov network will predict the wrong sign of the
interaction with 95% confidence.  The null model makes such errors more than 5 times
as often as the Markov network, even though it only reject the null hypothesis
twice as often overall (Appendix 4).  The Markov network's errors were also
more concentrated around 0, so it never misclassified strong interactions like
the null model did.

@@

<!--references go here-->

\setlength{\parindent}{0cm}
\setlength{\leftskip}{0cm}

\noindent \textbf{Figures:}

**Figure 1: A.** A small network of three competing species.
The tree (top) tends not to co-occur with either of the two shrub species, as
indicated by the strongly negative coefficient linking them. The two shrub
species also compete with one another, but more weakly (circled coefficient).
**B.** In spite of the competitive interactions between the two shrub species,
their shared tendency to occur in locations without trees makes their occurrence
vectors positively correlated (circled). **C.** Controlling for trees with a
conditional (all-else-equal) approach such as a partial covariance or a Markov
network leads to correct identification of the negative shrub-shrub interaction
(circled).

>>>>

**Abstract:** Inferring species interactions from co-occurrence data is one of
the most controversial tasks in community ecology. One difficulty is that a
single pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, the negative correlation between
two competing species can be reversed in the presence of a third species that is
capable of outcompeting both of them. Here, I apply models from statistical
physics, called Markov networks or Markov random fields, that can predict
the direct and indirect consequences of any possible species interaction matrix.
Interactions in these models can be estimated from observed co-occurrence
rates via maximum likelihood, controlling for indirect effects. Using simulated
landscapes with known pairwise interaction strengths, I evaluated Markov
networks and six existing approaches. The Markov networks consistently
outperformed other methods, correctly isolating direct interactions between
species pairs even when indirect interactions or abiotic factors largely
overpowered them. Two computationally efficient approximations, based on
controlling for indirect effects with linear or generalized linear models, also
performed well. Indirect effects reliably caused a common null modeling approach
to produce incorrect inferences, however.

@@

\noindent
To the extent that nontrophic species interactions (such as competition) affect
community assembly, ecologists might expect to find signatures of these
interactions in species composition data [@macarthur_population_1958;
@diamond_island_1975]. Despite decades of work and several major controversies,
however [@lewin_santa_1983; @strong_ecological_1984; @connor_checkered_2013],
existing methods for detecting competition's effects on community structure are
unreliable [@gotelli_empirical_2009]. In particular, species' effects on one
another can become lost in a web of indirect effects. For example, the
competitive interaction between the two shrub species in Figure 1A is
obscured by their shared tendency to occur in unshaded areas (Figure 1B). While
ecologists have long known that indirect effects can overwhelm direct ones 
[@levine_competitive_1976], most methods for drawing inferences from 
co-occurrence data do not control for these effects [e.g. @diamond_island_1975;
@strong_ecological_1984; @gotelli_empirical_2009; @veech_probabilistic_2013;
@pollock_understanding_2014]. As a result, ecologists do not have tools that allow
them to isolate direct interactions from indirect effects.

While competition doesn't reliably reduce co-occurrence rates at the
whole-landscape level (as most methods assume), it does still leave a
signal in the data (Figure 1C). After controlling for the presence
of the tree species (e.g. by splitting the data set into shaded and unshaded 
sites or by using a model that estimates conditional relationships amongs 
species), the two shrubs do have a negative association, with fewer co-occurrences 
in unshaded areas than would be expected if they occurred independently of one
another.

Following @azaele_inferring_2010, this paper shows that Markov networks
[undirected graphical models also known as Markov random fields;
@murphy_machine_2012] can provide a framework for understanding the
landscape-level consequences of pairwise species interactions,
and for estimating them from observed presence-absence matrices. Markov networks
have been used in diverse scientific fields in similar contexts for decades, from
physics [where nearby particles interact magnetically; @cipra_introduction_1987]
to spatial statistics [where adjacent grid cells have correlated values;
@harris_contact_1974; @gelfand_modelling_2005]. While community ecologists
explored some related approaches in the 1980's [@whittam_species_1981],
they used severe approximations that led to unintelligible results
[e.g. "probabilities" greater than one; @gilpin_factors_1982].

Below, I demonstrate Markov networks' ability to produce exact predictions
about the direct and indirect consequences of an interaction matrix, and also to
make inferences about the interaction matrix based on co-occurrence rates. Using 
simulated data sets where the "true" interactions are known, I compare this 
approach with several existing methods. Finally, I
discuss opportunities for extending the approach presented here to other
problems in community ecology, e.g. quantifying the overall effect of species
interactions on occurrence rates [@roughgarden_competition_1983] and
disentangling the effects of biotic versus abiotic interactions on species
composition [@pollock_understanding_2014].

@@

\noindent where the second sum is over all $\frac{1}{2}n(n-1)$ pairs of $n$
species. In this model, $\alpha_{i}$ is an intercept term determining the amount
that the presence of species $i$ contributes to the log-probability of
$\vec{y}$; it directly controls the prevalence of species $i$. Similarly,
$\beta_{ij}$ is the amount that the co-occurrence of species $i$ and species $j$
contributes to the log-probability; it controls the conditional relationship
between two species, i.e. the probability that they will be found together,
after controlling for the other species in the network (Figure 2A, Figure 2B).
For example, if $\beta_{ij} = +2$, then each species' odds of occurrence would be
$e^2$ times higher when the other one is present (as compared with otherwise
equivalent sites). The relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur. As a result, the model tends---all else
equal---to produce assemblages where many positively-associated species pairs
co-occur and few negatively-associated pairs do (just as an ecologist might
expect). When all else is *not* equal (e.g. Figure 1, where the presence of one
competitor is associated with release from another competitor), then predicting
species' overall co-occurrence rates would require normalizing the relative 
probabilities of possible $\vec{y}$ vectors so that they sum to one (Figure 2B).

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a set of independent $\vec{y}$ vectors
indicating which species are present at each site on the landscape). When
Equation 1 can be normalized, one can find exact maximum likelihood estimates for
$\alpha$ and $\beta$ using numerical optimization. Fully-observed networks like the 
ones considered here have unimodal likelihood surfaces [@murphy_machine_2012], so 
these estimates will represent the global optimum. I used the rosalia package 
[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to 
calculate $p(\vec{y}; \alpha, \beta)$ and its gradients [@murphy_machine_2012]; 
the package passes these functions to the "BFGS" method in R's general-purpose 
optimizer, which then estimates the Markov network parameters.

@@

I then simulated landscapes with up to 20 interacting species at 25, 200, or
1600 sites using three increasingly complex models (50 replicates for each
combination of size and model). The simplest set of simulated landscapes were
generated with Gibbs samples from Equation 1, and provided a test of methods'
ability to isolate direct species-level interactions from a web of indirect 
ones. For each replicate, I randomly drew the “true” $\beta$ coefficient
magnitudes an exponential distribution with rate 1 so that
most species pairs interacted negligibly but a few interactions were strong
enough to propagate through the newtwork. I randomly assigned 25% of the
interactions to be positive; the remainder were negative.

The next set of landscapes provided a way to assess each method's ability to 
identify direct interactions in the presence of environmental heterogeneity.
Here, the $\beta$ coefficients were calculated as above, but each species'
$\alpha$ value depended linearly on two normally-distributed environmental 
factors, which were drawn from independent Gaussians for each site. Once the 
local $\alpha$ values were calculated, independent Gibbs samplers determined 
the species composition for each site based on Equation 1.

In the third set of landscapes, I simulated each species' abundance (instead of
just their presence/absence), and interactions between species occurred on a 
per-capita basis. To avoid runaway mutualisms leading to infinite abundance,
all interaction coefficients were negative in these simulations. If a method
performs well on these landscapes, it would indicate some robustness to the
biolocical mechanisms underlying species' interactiosn with one another.

@@

I also evaluated six alternative methods: five from the existing literature,
plus a novel combination of two of these methods. The first alternative 
interaction metric was the sample correlation between species' presence-absence
vectors, which summarizes their marginal association.  Next, I used partial 
correlations, which summarize species' conditional relationships. This approach, 
which is  closely related to linear regression, is common in molecular biology
[@friedman_sparse_2008], but is rare in ecology (see @albrecht_spatial_2001 and
@faisal_inferring_2010 for two exceptions). In the context of non-Gaussian data,
the partial correlation can be thought of as a computationally efficient
approximation to the full Markov network model [@loh_structure_2013]. Partial
correlations are undefined for landscapes with perfectly-correlated species
pairs, so I used the regularized estimate provided by the corpcor package’s 
`pcor.shrink` function with the default settings [@schafer_corpcor_2014].

@@

The next method used the Pairs software described in @gotelli_empirical_2009.
This program simulating new landscapes from a null model that retains the row 
and column sums of the original matrix [@strong_ecological_1984] and calculates
$Z$-scores to summarize a species pair's deviation from this null.

@@

For the larger landscapes, I evaluated the relationship between each method's
estimates and the "true" interaction strengths. To ensure that the different
test statistics (e.g. correlations versus $Z$ scores) were on a common scale, I
rescaled them using linear regression through the origin. I then calculated the
proportion of variance explained for different combinations of model type and
landscape size (compared with a baseline model that assumed all interaction
strengths to be zero).

For the null model and the Markov network, the probability of rejecting the
null hypothesis of zero interaction was estimated across a range of "true"
interaction strengths using a generalized additive model with binomial responses
(reject versus not; Appendix 4). The Type I error rate was defined as this
model's estimated probability of rejecting the null hypothesis when the "true"
interaction was zero.

@@

\noindent
\textbf{Twenty species.}
In general, each model's performance was highest for large landscapes with
simple assembly rules and no environmental heterogeneity (Figure 3).
Despite some variability across contexts, the rank ordering across methods was
very consistent.  In particular, the four methods that controlled for indirect
effects (the Markov network, the generalized linear models, and the two partial
correlation-based methods) always matched or outperformed those that did not.
The Markov network consistently performed best of all. As anticipated by
@lee_learning_2012, generalized linear models closely
approximated the Markov network estimates (Figure 4A), especially when the data
sets were very large (Figure 3). As reviewed in @gotelli_empirical_2009,
however, most analyses in this field of ecology involve fewer than 50 sites;
in this context, the gap between the methods was larger. As shown in Appendix 4,
the standard errors associated with the estimates in Figure 3 are small (less
than 0.01), so the differences among methods should not be attributed to 
sampling error.

Of the methods that did not control for indirect effects, Figure 3 shows that
simple correlation coefficients provided a more reliable indicator of species'
true interaction strengths than either the joint species distribution model
(BayesComm) or the null model (Pairs). 95% of the variance the Pairs test 
statistic was explained by correlation coefficients (controlling for landscape
size; Figure 4B); much of the remaining variance is due to sampling error.

Finally, we can evaluate the models' statistical inferences (focusing on the
first two simulation types, where the interaction coefficients are easiest to
interpret).  The Markov network's Type I error rate depended on the 
simulation type: [[0.02]] for simulations that matched the model's assumptions,
versus [[0.14]] for simulations that included environmental heterogeneity (see 
Appendix 4 for confidence interval coverage across a range of $\beta_{ij}$ 
values). The null model's Type I error rates were [[0.30]] and [[0.50]] for the
constant and heterogeneous landscapes, respectively---far higher than the 
nominal 0.05 rate. Figure 4C shows, across a range of true interaction 
strengths, the probability that the null model or the Markov network will 
predict the wrong sign of the interaction with 95% confidence.  The null model 
makes such errors more than [[5]] times as often as the Markov network, even
though it only rejects the null hypothesis [[twice]] as often overall (Appendix
4). The Markov network's errors were also more concentrated around 0, so it 
never misclassified strong interactions like the null model did.

@@

**Figure 1: A.** A small network of three competing species.
The tree (top) tends not to co-occur with either of the two shrub species, as
indicated by the strongly negative coefficient linking them. The two shrub
species also compete with one another, but more weakly (circled coefficient).
**B.** In spite of the competitive interactions between the two shrub species,
their shared tendency to occur in locations without trees makes their occurrence
vectors positively correlated (circled). **C.** Controlling for trees with a
conditional (all-else-equal) approach such as a partial covariance or a Markov
network leads to correct identification of the negative shrub-shrub interaction
(circled). See Appendix 1 and the results for "three species" for more details.
